Compositional Quantitative Resource Analysis of C Programs

### 2014-06-26 paper/main.tex
typo
revamp automatic analysis a bit

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume that a program state
$\state$ simply maps variable to integers then we require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

@@

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplus$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state + P - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Tick}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q {+} M_t(n) }
  { \code{tick}(n) }
  { Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \mathbb Q_0^+
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

@@

The non-trivial part of this derivation is finding the loop
invariant that for the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximums so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
of the same program.  It can be done either by adapting the proof or by
applying the {\sc L:Frame} rule one top of the triple already derived.
More generally, it is clear that without function calls, the {\sc L:Frame}
rule is admissible in our system.

@@

\section{Automatic Amortized Analysis}

We define in this section an abstract version of the proof system that
we will use to derive bounds in our practical implementation.

\subsection{Linear Potential Functions}

To get a practical and automated implementation of amortized analysis
as presented in the overview section we need the fix the shape of
potential functions.  The general form of a potential function is as follows
$$
\Phi(H) = q_0 + \sum_{a \prec b} q_{ab}\,|H(a) - H(b)|.
$$
Where $a$ and $b$ range over the set of all program variables $V$ and
$(q_{ab})$ and $q_0$ are non-negative rational numbers.  The relation
$\prec$ is an arbitrary complete order on $V$ that is used only to
avoid considering the two equal quantities $|H(a) - H(b)|$ and
$|H(b) - H(a)|$ in the above sum.

To simplify the references to the $q$ coefficients later we introduce
an \emph{index set} $I$.  This set is defined to be $\{0\} \cup
\{(a, b) \mid a \prec b\}$.  Each index $i$ corresponds to a \emph{base
function} $f_i$ in the potential function: $0$ corresponds to the constant
function $\_ \mapsto 1$, and $(a,b)$ corresponds to $H \mapsto
|H(a) - H(b)|$.  Using these notations we can rewrite the above
equality as
$$
\Phi(H) = \sum_{i \in I} q_i f_i(H).
$$
From now on, we will write $ab$ to denote the index $(a,b)$.

\subsection{Proof Rules}

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
%
\\ \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \mathbb Q_0^+
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule{Q:IncP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:IncM}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Inc}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
\\\\ M = M_u + M_e(x {\pm} y)
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \mathbb Q_0^+
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\and \Rule{Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:proof}
\end{figure*}

The deductive system defined in~\pref{fig:proof} derives
judgements of the form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
Its semantics are formally described in the soundness section but we
will give an intuition of its meaning here.

\paragraph{Quantitative part.}
The judgement is really two parts, one logical part, and one quantitative
part.  The quantitative part is, as shown in the overview section, a
representation of two potential functions $\Phi$ and $\Phi'$ using
two maps from indices to non-negative rational numbers
$Q, Q' : I \rightarrow \mathbb Q_0^+$, these maps are called
\emph{quantitative contexts}.
We note $q_i$ for $Q(i)$.  Since $Q$ specifies a value for each
index in $I$ it is clear that it defines a unique linear potential function
as described in the previous section. We define $H \mapsto \Phi_H(Q)$
to be the linear potential function canonically associated to $Q$.

The rules of~\pref{fig:proof} are designed to enforce a semantic
soundness similar to the one exposed in the overview section.  To
do this, the rules impose contraints on $Q$ and $Q'$.  To make the
presentation lighter, we use the following  set of conventions.
First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
and $\forall i \neq 0.\, q'_i = q_i$.  Second, for any index $i$ such that
the relation between $q_i$ and $q'_i$ is not specified as a side condition
of a rule, we assume that $q_i = q'_i$.  Third, we implicitely assume
all the side conditions enforcing that the coefficients $(q_i)_{i \in I}$ are
non-negative numbers. Finally, when we write $q_x$ for $q_{x0}$ where
0 is a program variable symbolizing the constant 0.

\paragraph{Logical part.}
The second part of judgements maintains an abstract program state.
This should be seen as a very weak Hoare logic.  It has to be weak because
we want to automate it trivially and it can be weak because we only use
it to get very local and simple knowledge.  The specifics of logical assertions
in $\Gamma$ are left to the implementation section.
%
The purpose of this logical state is to help the rules
make some sound decisions when relating coefficients of quantitative
contexts.

Syntactically $\Gamma$ is a list of program conditions.  Because of the
rules {\sc Q:While} and {\sc Q:If}, we require the program conditions
to be closed by negation.  Semantically we define the validity
$H \models \Gamma$ of $\Gamma = C_1, \dots, C_n$ on a given
heap $H$ to be $\bigwedge_i \llbracket C_i \rrbracket_H = \top$.
%
The logical state
is maintained using regular Hoare style rules.
The semantic entailment on logical states is noted $\Gamma \models \Delta$,
its meaning is that $\forall H.\, H\models\Gamma \implies H\models\Delta$.

%\paragraph{Explanation of selected rules.}

>>>>

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} (LP) to
compute a derivation in the logic.  If we assume that a program state
$\state$ simply maps variable to integers then we require
$$
\Phi(\state) = q_0 + \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplusz$ and $|[a,b]| = \max(b-a,0)$.  Note that
$q_0$ is the constant potential and that we treat constants as program
variables.  For instance, we always have $q_{(0,x)}|[0,x]|$ as a
component of $\Phi(\state)$ if $x \in \dom{\state}$.
%
We then develop an inference rule for each syntactic construct that
derives sound triples $\htriple{\Phi}P{\Phi'}$ in the quantitative
logic but enables inference by using an LP solver.

@@

\pref{lem:sem1} summarizes the main properties of the resource
counter of the semantics.  It can be proved by induction on the number
of steps.
\begin{lemma}
\label{lem:sem1}
  \begin{enumerate}
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $n\ge 0$ then $(\state, S, \cont, \cost {+} n)
    \smallstep{M}^k (\state', S', \cont', \cost' {+} n)$.
  \item If $(\state, S, \cont, \cost) \smallstep{M}^k (\state', S',
    \cont', \cost')$ and $(\state, S, \cont, d) \smallstep{M}^k (\state', S',
    \cont', d')$ then $\cost - \cost' = d - d'$.
  \end{enumerate}
\end{lemma}
%
If we assume that we have a \emph{positive resource metric}, that is,
a metric that assigns a positive cost to each evaluation step then we
can link bounds on the resource consumption under the metric to
termination.  This is formalized by \pref{lem:sem_term}.
%
\begin{lemma}
 \label{lem:sem_term}
 Let $M$ be a positive metric and assume that there exists $b \in
 \Qplusz$ such that $(\state, S, \cont, \cost) \smallstep{M}^*
 (\state', S', \cont', \cost') \implies \cost {-} \cost' {\leq} b$.
 Then there exists $B \in \N$ such that $(\state, S, \cont, d)
 \smallstep{M}^n (\state'', S', \cont', d')$ for some $n$, $d$, $d'$
 and $\state''$ implies $n \le B$.
\end{lemma}

@@

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{L:Assert}
{ }
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state \implies Q + M_a }
  { \code{assert}~e }
  { Q }
}
%
\and
\Rule{L:Skip}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q }{ \code{skip} }{ Q }
}
%
\and
\Rule{L:Break}
{ }
{ \Delta; B; R \vdash
\htriple
  { M_b + B }{ \code{break} }{ Q }
}
%
\and
\Rule{L:Return}
{ }
{ \Delta; B; R \vdash
\htriple
  { R\,(\state(x)) }{ \code{return}~x }{ Q }
}
%
\and
\Rule{L:Update}
{ }
{ \Delta; B; R \vdash
\htriple
  { \lambda \state.\,M_u + Q\,\state[x \mapsto \sem{e}_\state] }{ x \gets e }{ Q }
}
%
\and
\Rule{L:Call}
{
  \Delta(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v) \\
  P \models P_f \, y \, (\state(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \state[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q - M_r }
}
%
\and
\Rule{L:Loop}
{ \Delta; Q; R \vdash
\htriple
  { I }{ S }{ I + M_l }
}
{ \Delta; B; R \vdash
\htriple
  { I }{ \code{loop}~S }{ Q }
}
%
\and
\Rule{L:Seq}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1 }{ Q' + M_s }
\\
  \Delta; B; R \vdash
\htriple{ Q' }{ S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S_1;S_2 }{ Q }
}
%
\and
\Rule{L:If}
{ \Delta; B; R \vdash
\htriple
  { \code{istrue}~\sem{e}_\state + P - M_c^1}
  { S_1 }{ Q }
\\
  \Delta; B; R \vdash
\htriple
  { \code{isfalse}~\sem{e}_\state + P - M_c^2 }
  { S_2 }{ Q }
}
{ \Delta; B; R \vdash
\htriple
  { P }{ \code{if}(e)~S_1~\code{else}~S_2 }{ Q }
}
%
\and
\Rule{L:Tick}
{ }
{ \Delta; B; R \vdash
\htriple
  { Q {+} M_t(n) }
  { \code{tick}(n) }
  { Q }
}
%
\and
\Rule{L:Weaken}
{
  P \models P' \\
  \Delta; B'; R' \vdash
  \htriple{ P' }{ S }{ Q' } \\
  Q' \models Q \\
  B' \models B \\
  \forall v.\, (R'\,v \models R\,v)
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\and
\Rule{L:Frame}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q } \\
  x \in \Qplusz
}
{ \Delta; B + x; R + x \vdash
\htriple{ P + x }{ S }{ Q + x }
}
%
\and
\Rule{L:Extend}
{ \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
\\
  \forall f\,P_f\,Q_f.\,
  \Delta'(f) = \forall z\,\vec v\,v.(P_f\,z\,\vec v, Q_f\,z\,v)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ S_f }{ \bot })
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}
%
\end{mathpar}

@@

The non-trivial part of this derivation is finding the loop
invariant for the while loop. To this end
it is necessary to define the number $A(i)$ of ``maximums so far
starting from $i$ in the array \lstinline{a}.''  When the condition
\lstinline{a[i] > m} is true, we know that we encountered a ``maximum
so far'' because $m$ is a maximal element of \lstinline{a[$0 \dots i$]},
thus $A(i) = 1 + A(i+1)$ and we get one potential unit to pay for the
assignment.  In the other case, no maximum so far is encountered so
$A(i) = A(i+1)$.  As a side remark, if $K \ge 0$, it is possible to
have $\{ A(0) + N + K \}$ and $\{ K \}$ as pre- and postcondition
of the same program.  It can be done either by adapting the proof or by
applying the {\sc L:Frame} rule one top of the triple already derived.
More generally, it is clear that without function calls, the {\sc L:Frame}
rule is admissible in our system.

@@

\section{Automatic Analysis}

In this section we present the deductive system used in our practical
implementation to derive linear bounds on C programs.

\paragraph{Linear Potential Functions.}

To find derivations in our Quantitative Logic automatically, the main
insight is to consider only assertions that match a certain shape.
The general form of a potential function that we consider is
$$
  \Phi(\state) = q_0 + \sum_{x, y \in \dom\state}
    q_{(x, y)} \cdot |\inter {\state(x)} {\state(y)}|.
$$

To simplify the references to the linear coefficients $q$, we introduce
an \emph{index set} $I$.  This set is defined to be $\{0\} \cup
\{(x, y) \mid x, y \in \Var \}$.  Each index $i$ corresponds to a \emph{base
function} $f_i$ in the potential function: $0$ corresponds to the constant
function $\_ \mapsto 1$, and $(x,y)$ corresponds to $\state \mapsto
|\inter {\state(x)} {\state(y)}|$.  Using these notations we can rewrite the above
equality as
$$
  \Phi = \sum_{i \in I} q_i f_i.
$$
We will often write $xy$ to denote the index $(x,y)$.  The family
$(f_i)_{i\in I}$ is actually a basis (in the linear algebra sense)
for potential functions, this allows us to represent uniquely any
linear potential function $\Phi$ as a \emph{quantitative annotation}
$Q = (q_i)_{i \in I}$, that is a family of rational numbers where
only a finite number of elements are not zero.
%
For soundness, we require that potential functions always give
non-negative results.  This is acheived by restricting coefficients
in a quantitative annotation to be always non-negative.

\paragraph{Logical State.}

Quantitative annotations alone are not enough to make our system
functional.  A minimal logical state needs to be maintained to
justify certain operations on quantitative annotations.  For
example when analyzing the code $x \gets x + y$, it is helpful
to know the sign of $y$ to determine which intervals will increase
or decrease.  The knowledge needed by our rules can often be
inferred by local reasoning within usual theories (e.g. Presburger
arithmetic or bit vectors).

Our implementation uses conjunctions of linear inequalities as
logical state.  We never compute fixpoints and take $\top$
as pre- and postconditions for functions. It has proved to be
sufficient for the variety of examples we designed and found
in the literature.

In contrast to the quantitative annotations, logical contexts in
the presented system are left abstract.  This allows for simpler
rules and leaves room for numerous improvements on top of our
minimal implementation.

\paragraph{Deduction Rules.}

The deductive system defined in~\pref{fig:auto} derives
judgements of the form $\Gamma; Q \vdash S \dashv Q'; \Gamma'$.
Its semantics are formally described in the soundness section but we
will give an intuition of its meaning here.

\paragraph{Quantitative part.}
The judgement is really two parts, one logical part, and one quantitative
part.  The quantitative part is, as shown in the overview section, a
representation of two potential functions $\Phi$ and $\Phi'$ using
two maps from indices to non-negative rational numbers
$Q, Q' : I \rightarrow \Qplusz$, these maps are called
\emph{quantitative contexts}.
We note $q_i$ for $Q(i)$.  Since $Q$ specifies a value for each
index in $I$ it is clear that it defines a unique linear potential function
as described in the previous section. We define $H \mapsto \Phi_H(Q)$
to be the linear potential function canonically associated to $Q$.

The rules of~\pref{fig:auto} are designed to enforce a semantic
soundness similar to the one exposed in the overview section.  To
do this, the rules impose contraints on $Q$ and $Q'$.  To make the
presentation lighter, we use the following  set of conventions.
First, the context $Q'$ noted $Q \pm n$ is defined by $q'_0 = q_0 \pm n$
and $\forall i \neq 0.\, q'_i = q_i$.  Second, for any index $i$ such that
the relation between $q_i$ and $q'_i$ is not specified as a side condition
of a rule, we assume that $q_i = q'_i$.  Third, we implicitely assume
all the side conditions enforcing that the coefficients $(q_i)_{i \in I}$ are
non-negative numbers. Finally, when we write $q_x$ for $q_{x0}$ where
0 is a program variable symbolizing the constant 0.

\begin{figure*}[t]
\begin{mathpar}
%
\Rule{Q:Skip}
{}
{ B; R; (\Gamma, Q) \vdash \code{skip} \dashv (\Gamma, Q) }
%
\and \Rule{Q:Break}
{}
{  (\Gamma, Q_B); R; (\Gamma, Q_B {+} M_b) \vdash \code{break} \dashv (\Gamma', Q') }
%
\and \Rule{Q:Return}
{ P = Q_R[\Vret/x]  
\\ \forall i \in \dom{P} . \, p_i = q_i
}
{ B; (\Gamma,Q_R); (\Gamma, Q{+} M_r) \vdash \code{return}~x \dashv (\Gamma',Q') }
%
\and \Rule{Q:Assert}
{}
{ B; R; (\Gamma, Q {+} M_a) \vdash \code{assert}~e \dashv (\Gamma \land e, Q) }
%
\and \Rule{Q:Tick}
{}
{ B; R; (\Gamma, Q{+}M_t(n)) \vdash \code{tick}(n) \dashv (\Gamma, Q) }
%
\\ \Rule{Q:Loop}
{ (\Gamma', Q'); R; (\Gamma,Q) \vdash S \dashv (\Gamma,  Q {+} M_l) }
{ B; R; (\Gamma, Q) \vdash \code{loop}~S \dashv (\Gamma', Q') }
%
\and \Rule{Q:Update}
{ q'_{xy}, q'_{yx} \in \Qplusz
\\ \forall u. (q_{yu} = q'_{xu} + q'_{yu} \land q_{uy} = q'_{ux} + q'_{uy})
}
{ B; R; (\Gamma[x/y], Q {+} M_u {+} M_e(y)) \vdash x \gets y \dashv (\Gamma, Q') }
%
\and \Rule{Q:IncP}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x + y \in \inter x u \right\}
\\ \textstyle q'_{0y} = q_{0y}
      + \sum_{u \in \mathcal U} q_{xu}
      - \sum_{v \not\in \mathcal U} q_{vx}
}
{ \textstyle
  B; R; (\Gamma[x/x {+} y], Q {+} M_u {+} M_e(x {+} y)) \vdash x \gets x + y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:IncM}
{ \Gamma \models 0 \le y
\\ \mathcal U = \left\{ u \mid \Gamma \models x - y \in \inter u x \right\}
\\ \textstyle q'_{y0} = q_{y0}
      + \sum_{u \in \mathcal U} q_{ux}
      - \sum_{v \not\in \mathcal U} q_{xv}
}
{ \textstyle
  B; R; (\Gamma[x/x {-} y],Q {+} M_u {+} M_e(x {-} y)) \vdash x \gets x - y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:Inc}
{ \textstyle q'_{0y} = q_{0y} - \sum_{v} q_{vx}
\\ \textstyle q'_{y0} = q_{y0} - \sum_{v} q_{xv}
\\\\ M = M_u + M_e(x {\pm} y)
}
{ \textstyle
  B; R; (\Gamma[x/x {\pm} y], Q {+} M) \vdash x \gets x \pm y \dashv (\Gamma, Q')
}
%
\and \Rule{Q:If}
{ B; R; (\Gamma \land e, Q {-} M_c^1) \vdash S_1 \dashv (\Gamma',Q')
\\\\ B; R; (\Gamma \land \neg e, Q {-} M_c^2) \vdash S_2 \dashv (\Gamma',Q')
}
{ B; R; (\Gamma, Q) \vdash \code{if}(e)~S_1~\code{else}~S_2 \dashv (\Gamma',Q') }
%
\and\Rule{Q:Seq}
{  B; R; (\Gamma, Q) \vdash S_1 \dashv (\Gamma', Q' {+} M_s)
\\\\ B; R; (\Gamma', Q') \vdash S_2 \dashv (\Gamma'',Q'')
}
{ B; R; (\Gamma, Q) \vdash S_1;S_2 \dashv (\Gamma'',Q'') }
%
\and \Rule{Q:Weak}
{ B; R; (\Gamma_2,Q_2) \vdash S \dashv (\Gamma'_2, Q'_2)
\\\\ \Gamma_1 \models \Gamma_2
\\ Q_1 \succeq_{\Gamma_1} Q_2
\\ Q'_2 \succeq_{\Gamma_2'} Q'_1
\\ \Gamma'_2 \models \Gamma'_1
}
{ B; R; (\Gamma_1,Q_1) \vdash S \dashv (\Gamma'_1, Q'_1) }
%
\and\Rule{Q:Call}
{ (\Gamma_f, Q_f,  \Gamma_f', Q_f') \in \Delta(f)
\and \text{Loc} = \text{Locals}(Q)
\and \forall i \neq j . \, x_i \neq x_j
\and c \in \Qplusz
\\ Q = P + S
\and Q' = P' + S
\and U = Q_f[\Vargs / \vec x]
\and U' = Q'_f[\Vret/r]
\\ \forall i \in \dom{U} . \, p_i = u_i
\and \forall i \in \dom{U'} .\, p'_i = u'_i 
\and \forall i \not\in \dom{U'} .\, p'_i = 0
\and \forall i \not\in \text{Loc}
  .\, s_i = 0
}
{ B; R; (\Gamma_f[\Vargs / \vec x] \land \Gamma_{\text{Loc}}, Q {+} c {+} M_f)
  \vdash r \gets f(\vec x) \dashv
  (\Gamma_f'[\Vret/r] \land \Gamma_{\text{Loc}}, Q' {+} c {-} M_r)
}
%
\and \Rule{Relax}
{ \mathcal L = \{ xy \mid \exists l_{xy},~\Gamma \models l_{xy} \le |\inter x y| \}
\and \mathcal U = \{ xy \mid \exists u_{xy},~\Gamma \models |\inter x y| \le u_{xy} \}
\\ \forall i \in \mathcal U.\, q'_i \ge q_i - p_i
\\ \forall i \in \mathcal L.\, q'_i \ge q_i + p_i
\\ \forall i \not\in \mathcal U \cup \mathcal L.\, q'_i \ge q_i
}
{ Q' + \sum_{i \in \mathcal U} u_i p_i - \sum_{i \in \mathcal L} l_i p_i  \succeq_\Gamma Q }
%
\and \Rule{Extend}
{ \fenv f = (\vec y, S_f)
\\ B; (\Gamma_f', Q'_f); (\Gamma_f[\Vargs/\vec y], Q_f[\Vargs/ \vec y] ) \vdash S \dashv (\Gamma', Q')
}
{ (\Gamma_f, Q_f, \Gamma_f', Q_f') \in \Delta(f) }
%
\end{mathpar}
\caption{Inference rules of the quantitative analysis}
\label{fig:auto}
\end{figure*}
